Robust PCA and Higher Moments Distances for Botnet Detection

### 2019-04-20 docs/papers/2019JNCA_m-rpca/2019JNCA_m-rpca.tex
adding initial paper template

\title{Cumulant-based Robust Estimates for Network Attack Detection}

%% Group authors per affiliation:
\author[unbaddress]{Thiago P. B. Vieira}
\author[unbaddress,Ilmenauaddress,Fraunhoferaddress]{João Paulo C. L. da Costa}
\author[Ilmenauaddress,Fraunhoferaddress]{Giovanni Del Galdo}

@@

\begin{keyword}
Network Attack Detection \sep Cumulant \sep Minimum Covariance Determinant \sep Mahalanobis Distance
\end{keyword}

@@

Traditionally, cyber defense methods can be effective against ordinary and conventional types of attacks, yet may fail against innovative malicious techniques \cite{lakhina2005mining}. In order to be able to detect and avoid novel attacks and their variations, it is necessary to develop or improve techniques to achieve efficiency on resource consumption, processing capacity and response time. Moreover, it is crucial to obtain high detection accuracy and capacity to detect variations of malicious patterns. Recently, signal processing schemes have being applied to the detection of malicious traffic in computer networks \cite{Lu2009,Huang2009,Zonglin2009,david2011blind,da2012improved,tenorio2013greatest}, showing advances in network traffic analysis.

Information security may consist of both technical and procedural aspects. The former includes equipment and security systems, while the latter corresponds to security rules and recommendations. Intrusion detection and intrusion prevention systems are security systems used, respectively, to detect (passively) and prevent (proactively) threats to computer systems and computer networks. Such systems can work in the following fashions: signature-based, anomaly-based or hybrid \cite{Huang2009,mudzingwa2012study}. Additionally, anomaly detection techniques can be categorized in classification, statistical, information theory and clustering based, according to \cite{bhuyan2014network,ahmed2016survey,osanaiye2016distributed}.

@@

@@

Several methods have been proposed for the identification and characterization of malicious activity in computer networks. Classical methods typically employ data mining \cite{he2008applying,ghourabi2010data,osanaiye2016distributed} and regular file analysis \cite{raynal2004honeypot} to detect patterns that indicate the presence of specific attacks in network traffic.

Data mining is often used to describe the process of extracting useful information from large databases. Multiple methods of data mining are used in \cite{he2008applying,osanaiye2016distributed} to analyze data flow in a network with the aim of identifying characteristics of malicious traffic in large scale environments. Researchers have applied data mining techniques in log analysis \cite{ghourabi2010data} to improve intrusion detection performance. However, data mining techniques used so far in network analysis require prior collection of large data sets, which is a limitation of several schemes for online analysis \cite{Huang2009}.

Regular file analysis \cite{raynal2004honeypot} consists of traffic analysis for detecting known patterns that indicate the presence of attacks, applying statistical analysis to the study of collected traffic. An essential feature of this method is that it depends on prior knowledge of the details of the attacks to be identified, and also depends on previous log collection for traffic analysis and false positives reduction.

@@

Besides being prone to higher errors and false positives, such human intervention makes PCA not useful for real time applications. Therefore, in order to automate the analysis of eigenvalues profile, model order selection (MOS) schemes should be incorporated.

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work requires a training method to produce a prediction model for normal daily traffic and presents limitations on identification of behaviors without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work is not applied to probe and flood attack detection, simultaneously.

The data collected from honeypot systems, such as captured traffic and operating system logs, can be analyzed to obtain information about attack techniques, general trends of threats and exploits. Blind automatic detection of malicious traffic techniques have been developed for honeypots in \cite{david2011blind,da2012improved}. However, traffic on honeypot is simpler than real network traffic, because there are no running legitimate applications, due to the fact that honeypots emulate behavior of a host within a network to deceive and lure attackers \cite{zakaria2012review}. Since honeypots do not generate legitimate traffic, the amount of data captured in honeypots is significantly lower in comparison to a Network Intrusion Detection System (NIDS), which captures and analyzes the largest possible amount of network traffic \cite{david2011blind}. MOS for blind identification of malicious activities in honeypots was proposed by us in \cite{david2011blind}, which evaluated criteria for selecting the model order, through simulations and comparing the order of the resulting model with the true model order.

The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probe and flood attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.

Several approaches for network attack detections uses the KDD 99 \cite{ji2016multi,ahmed2016survey,osanaiye2016distributed,bhuyan2014network} datasets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 dataset are criticized by for the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled datasets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} dataset is the refined version of the KDD 99 dataset that redundant data records are removed, in order to avoid biased classifications. Additionally, some approaches uses simulated \cite{callegari2011novel} scenarios or non-public datasets their evaluations. Since the proposed approach relies on a packet level analysis and the KDD 99 and NSL-KDD datasets adopt a traffic aggregation by connections, we consider the use of a experimental scenario on a real network and the DARPA 1998 dataset, which is the source for the creation of the KDD 99 and NSL-KDD datasets. Note that the proposed approach is not based on learning or classification techniques, which are more susceptible to biased results caused by the issues in the DARPA/KDD datasets.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cumulant-based Minimum Covariance Determinant}
\label{sec:c_mcd}

>>>>

\title{Robust PCA and Higher Moments Distances for Botnet Detection}

%% Group authors per affiliation:
\author[unbaddress]{Thiago P. B. Vieira}
\author[unbaddress]{Eduardo Said Calil Vilaça}
\author[unbaddress,Ilmenauaddress,Fraunhoferaddress]{João Paulo C. L. da Costa}
\author[unbaddress]{Rafael T. de Sousa Júnior}

@@

\begin{keyword}
Botnet Detection \sep Network Anomaly Detection \sep Skewness \sep Kurtosis \sep Robust Principal Component Analysis \sep Mahalanobis Distance
\end{keyword}

@@

@@

Anomaly detection is the identification of rare and suspicions observations by differing the normal or majority of the data.

Anomalies can be related to network attack, fraud, defect or medical problems, for examples.

Anomalies are also referred to as outliers, novelties, noise, deviations and exceptions.

Anomalies can be hard to identify and separate from normal data.

Considering that anomalies are rare in comparison to normal events, anomaly detection algorithms have to deal with problems imposed by imbalanced data.

@@

@@

Besides being prone to higher errors and false positives, such human intervention makes PCA not useful for real time applications. Therefore, in order to automate the analysis of eigenvalues profile, model order selection (MOS) schemes should be incorporated. Vieira .... (The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probe and flood attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.)

Several approaches for network attack detections uses the KDD 99 \cite{ji2016multi,ahmed2016survey,osanaiye2016distributed,bhuyan2014network} datasets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 dataset are criticized by for the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled datasets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} dataset is the refined version of the KDD 99 dataset that redundant data records are removed, in order to avoid biased classifications. Additionally, some approaches uses simulated \cite{callegari2011novel} scenarios or non-public datasets their evaluations. We propose to adopt CTU-13...

Our proposal is ...

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Robust PCA and Higher Moments Distances for Botnet Detection}
\label{sec:m_rpca}
